---
title: "Effects of Natural Disasters in USA From NOAA Storm Database, both Health and Economic Effects are Explored"
author: "SAI"
date: "July 22, 2018"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Synopsis
In united states we have large varieties of Natural Disasters that affect and cause damage both from economic point and 
health point. SO we are trying to analyze which type of natural disasters are causing more affects so that we can take necessary actions.

The data has been taken from the NOAA database and analysed to give the following two solutions:

1. Across the United States, which types of events (as indicated in the EVTYPE variable) are most harmful with respect to population health?
2. Across the United States, which types of events have the greatest economic consequences?

dplyr package is used to process the data and to produce the plots that can give us the perfect solutions to the above questions

## Data Processing
STEP1: Iputting the data in to RStudio and getting the feel of data including the dimensions
```{r StormData}
Input_data <- read.csv("repdata%2Fdata%2FStormData.csv.bz2",sep = ",", header = TRUE)
head(input)
dim(input)
names(input)
library(dplyr)
```

## Gathering The required data using the dplyr package

The dplyr package is used to gather the data regarding the natural disasters that are cauisng problems both economically and health point of view
```{r DataGathering, cache=TRUE}

Total_combineddata <- Input_data %>% select(EVTYPE, FATALITIES, INJURIES)
Fatalities_data <- Input_data %>% select(EVTYPE, FATALITIES)
Injuries_data   <- Input_data %>% select(EVTYPE, INJURIES)



```

Note that the `echo = FALSE` parameter was added to the code chunk to prevent printing of the R code that generated the plot.
